Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-25 paper/drinking.tex

\subsection{Processes, their participants and qualities}
\begin{figure*}
  \centering
  \includegraphics[width=.9\textwidth]{onto.pdf}
  \caption{Overview over ...\label{fig:onto}}
\end{figure*}

Processes are entities that are temporally extended: processes have
temporal parts and {\em unfold} through time. Examples of classes that
may have processes as instances include {\em Drinking}, {\em
  Triathlon} or {\em Apoptosis}.

@@

\section{Results}
\subsection{Attributes of processes}
An attribute that can be directly linked to a process is its {\em
  duration}. A duration is an attribute that characterizes the
temporal extent of a process and is similar to {\em Length}, {\em
  Area} and {\em Volume} for one-, two- and three-dimensional
entities. Furthermore, processes may be attributed a {\em temporal
  location} with respect to a particular coordinate system.

@@

\subsection{Cellular Phenotype Ontology}
A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth,
  height=.33\textheight]{overview.pdf}
  \caption{Overview over ...\label{fig:overview}}
\end{figure*}

@@

\subsection{Use of CPO}

The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

\subsection{Future research: acts and intentional states}
activities/actions, intentional (goal-oriented) acts

feeling thirsty (internal state) as driving force behind drinking
behavior

non-measurable, inference only

>>>>

\subsection{Processes, their participants and qualities}
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{onto.pdf}
  \caption{Overview over ...\label{fig:onto}}
\end{figure}

Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities. Examples of
classes that may have processes as instances include {\em Drinking},
{\em Triathlon} or {\em Apoptosis}.  Figure \ref{fig:onto} illustrates
these basic categories of being.

@@

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes:
process attributes that are linked directly to processes, attributes
that arise due to process parts and attributes of processes, and
attributes arising from attributes of process participants and other
process attributes.

Attribute that can be directly linked to a process are the result of
the temporal extension of a process and include process {\em duration}
and {\em temporal location}. A duration is an attribute that
characterizes the temporal extent of a process and is similar to {\em
  Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional entities. A {\em Temporal location} positions the
time interval during which the process occurs with respect to a
reference coordinate system.

@@

\subsection{Cellular Phenotype Ontology}
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth,
  height=.33\textheight]{overview.pdf}
  \caption{Overview over ...\label{fig:overview}}
\end{figure*}

@@

\subsection{Use of CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

% \subsection{Future research: acts and intentional states}
% activities/actions, intentional (goal-oriented) acts

% feeling thirsty (internal state) as driving force behind drinking
% behavior

% non-measurable, inference only
